Consistent ecosystem functional response across precipitation extremes in a sagebrush steppe

### 2017-07-26 manuscript/usses_water_ms.Rmd
addresses ARK minor comments

---
title: Consistent ecosystem functional response across precipitation extremes in a sagebrush steppe
preprint: false
author: 
  - name: Andrew T. Tredennick
    affiliation: 1
    corresponding: true
    email: atredenn@gmail.com
  - name: Andrew R. Kleinhesselink
    affiliation: 2
  - name: Bret Taylor
    affiliation: 3
  - name: Peter B. Adler
    affiliation: 1
affiliation:
  - code: 1
    address: Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah 84322
  - code: 2
    address: Department of Ecology and Evolutionary Biology, University of California, Los Angeles, Los Angeles, California 90095
  - code: 3
    address: USSES, address...
abstract: >
  Precipitation is predicted to become more variable in the western United States, meaning years of above and below average precipitation will become more common. Periods of drought and deluge are major drivers of ecosystem functioning in water limited grasslands, but how ecosystems respond to precipitation may change as the duration of above and below average periods increases. Changes in ecosystem functional response could reflect underlying changes in species composition or species reaching physiological thresholds at extreme precipitation levels. Here we report the results of a five-year experiment where we used drought and irrigation treatments (approximately 50% decrease/increase) to investigate whether ecosystem functional response remains consistent at extreme precipitation levels. We also use species-level adundance data to link ecosystem functional response to community dynamics. Ecosystem functional response, defined as the relationship between soil moisture and aboveground net primary productivity, was suprisingly consistent across drought, control, and irrigation treatments. Aboveground net primary productivity (ANPP) responded positively to availabile soil moisture, but the response was weak and did not vary across treatments. There was also no evidence that treatment effects grew over time. The similarity of ecosystem functional response across treatments was not due to compensatory shifts at the plant community level, where species composition among treatments was similar and remarkably stable over the five years. At least in the short-term, ecosystem functional response and community composition in this sagebrush steppe system is resistant to increases and decreases in growing season precipitation, perhaps due to bet-hedging strategies of constituent species.
header-includes: >
  \usepackage{setspace}
  \usepackage{todonotes}
  \usepackage{rotating}
  \usepackage{color, soul}
  \usepackage{sectsty}
  \usepackage{bm,mathrsfs}
bibliography: ~/Dropbox/Bibliography/usses_water.bib
output: 
  rticles::peerj_article:
    base_format: rmarkdown::pdf_document # bookdown::pdf_document2 # for using \@ref()
---

@@

At a given site, the functional response of ANPP to water availability (e.g., soil moisture) can be characterized by fitting a model to historical observations of ANPP and soil moisture.
However, the fitted functional response may provide an incomplete picture because future conditions are likely to be outside the historical range of variability [@Smith2011].
For example, historical trends may show a linear soil moisture-ANPP relationship, but that relationship could potentially saturate if soil moisture is pushed far beyond typical levels.
Saturating relationships are actually common [@Hsu2012], perhaps because other resources, like nitrogen, become more limiting in wet years than dry years.
Knowing the curvature of the soil moisture-ANPP relationship is critical for understanding how ecosystems will respond to chronic alterations in water availability.

@@

Manipulating potentially limiting resources offers a route to understanding how ecosystems will respond to resource levels that fall outside the historical range of variability [@Avolio2015; @Gherardi2015; @Knapp2017].
Altering the amount of precipitation over many years should provide insight into the time scales at which water-limited ecosystems respond to chronic resource alteration.
Following the HRF, we propose four alternative predictions for the effect of precipitation manipulation on the ecosystem functional response to soil moisture, that is, the soil moisture-ANPP relationship (Figure 1).
The four predictions are based on possible outcomes at the community (e.g., community composition) and ecosystem (e.g., soil moisture-ANPP regression) levels.

@@

All four outcomes are possible in any given ecosystem, but the time scales at which the different scenarios play out likely differ [@Smith2009; @Wilcox2016; @Knapp2017].
Thus, our task is not to test the validity of the HRF, but rather to amass information on how quickly ecosystem functional responses change in different ecosystems.
Likewise, we need to understand whether changes at the ecosystem level are driven community level changes or individual level responses.

@@

## Study Area
We conducted our precipitation manipulation experiment at the United States Sheep Experimental Station (USSES) near Dubois, Idaho (44.2$^{\circ}$ N, 112.1$^{\circ}$ W), 1500 m above sea level.
The vegetation is typical of high elevation sagebrush steppe.
The plant community is dominated by the shrub *Artemesia tripartita* and three perennial bunchgrasses, *Pseudoroegneria spicata*, *Poa secunda*, and *Hesperostipa comata*.
During the period of our experiment (2011 -- 2015), average mean annual precipitation was 265 mm year$\phantom{}^{-1}$ and mean monthly temperature ranged from -5.2$^{\circ}$C in January to 21.8$^{\circ}$C in July.

## Precipitation Experiment
Between 1926 and 1932, range scientists at the USSES established 26 permanent 1 m$^2$ quadrats to track vegetation change over time.
In 2007, we (well, one of us [P. Adler]) relocated 14 of the original quadrats, six of which were inside a large, permanent livestock exclosure.
We use these six plots as control plots that have recieved no treatment, just ambient precipitation.
In spring 2011, we (well, two of us [A. Kleinhesselink and P. Adler]) established 16 new 1 m$^2$ plots.
We avoided areas on steep hill slopes, areas with greater than 20\% cover of bare rock, and areas with greater than 10\% cover of the shrubs *Purshia tridentata* and/or *Amelanchier utahensis*.
We established the new plots in pairs and randomly assigned each plot in a pair to receive a "drought" or "irrigation" treatment.

@@

To make sure the treatments were having the desired effects, we monitored soil moisture in four of the drought-irrigation pairs using Decagon Devices (Pulman, Washington) 5TM and EC-5 soil moisture sensors.
We installed four sensors in each plot, two at 5 cm soil depth and two at 25 cm soil depth.
We also installed four sensors in areas nearby the four selected plot pairs to measure ambient soil moisture at the same depths.
Soil moisture measurements were automatically logged every four hours.
We coupled this temporally intensive soil moisture sampling with spatially extensive readings taken at six points within all 16 plots and associated ambient measurement areas.
These snapshot data were collected on 06/06/2012, 04/29/2015, 05/07/2015, 06/09/2015, and 05/10/2016 using a handheld EC-5 sensor.

\ark{ANDY: add paragraph on the statistical modeling of soil moisture to produce Figure 2B (same as your Figure 2 in the precip-demography-prediction dissertation chapter).}

@@

\begin{figure}[!ht]
  \centering
      \includegraphics[height=5in]{../figures/Figure1.png}
  \caption{(A) Probability density of historical precipitation from 1926-2016, with the years of the experiment shown with arrows on the \emph{x}-axis. (B) Observed soil volumetric water content (VWC) over the course of the experiment. (C) Mean (points) ANPP and its standard deviation (error bars) for each year of the experiment.}
\end{figure}

>>>>

---
title: Consistent ecosystem functional response across precipitation extremes in a sagebrush steppe
preprint: false
author: 
  - name: Andrew T. Tredennick
    affiliation: 1
    corresponding: true
    email: atredenn@gmail.com
  - name: Andrew R. Kleinhesselink
    affiliation: 2
  - name: Bret Taylor
    affiliation: 3
  - name: Peter B. Adler
    affiliation: 1
affiliation:
  - code: 1
    address: Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah 84322
  - code: 2
    address: Department of Ecology and Evolutionary Biology, University of California, Los Angeles, Los Angeles, California 90095
  - code: 3
    address: USSES, address...
abstract: >
  Precipitation is predicted to become more variable in the western United States, meaning years of above and below average precipitation will become more common. Periods of drought and deluge are major drivers of ecosystem functioning in water limited grasslands, but how ecosystems respond to precipitation may change as the duration of above and below average periods increases. Changes in ecosystem functional response could reflect compensatory changes in species composition or species reaching physiological thresholds at extreme precipitation levels. Here we report the results of a five-year experiment where we used drought and irrigation treatments (approximately 50% decrease/increase) to investigate whether ecosystem functional response remains consistent at extreme precipitation levels. We also use species-level adundance data to link ecosystem functional response to community dynamics. Ecosystem functional response, defined as the relationship between soil moisture and aboveground net primary productivity, was suprisingly consistent across drought, control, and irrigation treatments. Aboveground net primary productivity (ANPP) responded positively to availabile soil moisture, but the response was weak and did not vary across treatments. There was also no evidence that treatment effects grew over time. The similarity of ecosystem functional response across treatments was not due to compensatory shifts at the plant community level, where species composition among treatments was similar and remarkably stable over the five years. At least in the short-term, ecosystem functional response and community composition in this sagebrush steppe system is consistent, even at extreme precipitation levels. Such consistency may be due to bet-hedging strategies of constituent species.
header-includes: >
  \usepackage{setspace}
  \usepackage{todonotes}
  \usepackage{rotating}
  \usepackage{color, soul}
  \usepackage{sectsty}
  \usepackage{bm,mathrsfs}
bibliography: ~/Dropbox/Bibliography/usses_water.bib
output: 
  rticles::peerj_article:
    base_format: rmarkdown::pdf_document # bookdown::pdf_document2 # for using \@ref()
---

@@

At a given site, the functional response of ANPP to water availability (e.g., soil moisture) can be characterized by fitting a model to historical observations of ANPP and soil moisture.
However, the fitted functional response may provide an incomplete picture because future conditions are likely to be outside the historical range of variability [@Smith2011].
For example, historical trends may show a linear soil moisture-ANPP relationship, but that relationship could potentially saturate if soil moisture is pushed far beyond typical levels.
Saturating relationships are actually common [@Hsu2012; @Gherardi2015a], perhaps because other resources, like nitrogen, become more limiting in wet years than dry years.
Knowing the curvature of the soil moisture-ANPP relationship is critical for understanding how ecosystems will respond to chronic alterations in water availability.

@@

Manipulating potentially limiting resources like precipitation offers a route to understanding how ecosystems will respond to resource levels that fall outside the historical range of variability [@Avolio2015; @Gherardi2015; @Knapp2017].
Altering the amount of precipitation over many years should provide insight into the time scales at which water-limited ecosystems respond to chronic resource alteration.
Following the HRF, we propose four alternative predictions for the effect of precipitation manipulation on the ecosystem functional response to soil moisture, that is, the soil moisture-ANPP relationship (Figure 1).
The four predictions are based on possible outcomes at the community (e.g., community composition) and ecosystem (e.g., soil moisture-ANPP regression) levels.

@@

All four outcomes are possible in any given ecosystem, but the time scales at which the different scenarios play out likely differ [@Smith2009; @Wilcox2016; @Knapp2017].
Thus, our task is not to test the validity of the HRF, but rather to amass information on how quickly ecosystem functional responses change in different ecosystems.
Likewise, we need to understand whether changes at the ecosystem level are driven by community level changes or individual level responses.

@@

## Study Area
We conducted our precipitation manipulation experiment in a sagebrush steppe community at the United States Sheep Experimental Station (USSES) near Dubois, Idaho (44.2$^{\circ}$ N, 112.1$^{\circ}$ W), 1500 m above sea level.
The plant community is dominated by the shrub *Artemesia tripartita* and three perennial bunchgrasses, *Pseudoroegneria spicata*, *Poa secunda*, and *Hesperostipa comata*.
During the period of our experiment (2011 -- 2015), average mean annual precipitation was 265 mm year$\phantom{}^{-1}$ and mean monthly temperature ranged from -5.2$^{\circ}$C in January to 21.8$^{\circ}$C in July.

## Precipitation Experiment
Between 1926 and 1932, range scientists at the USSES established 26 permanent 1 m$^2$ quadrats to track vegetation change over time.
In 2007, we (well, one of us [P. Adler]) relocated 14 of the original quadrats, six of which were inside a large, permanent livestock exclosure.
We use these six plots as control plots that have recieved no treatment, just ambient precipitation.
In spring 2011, we (well, two of us [A. Kleinhesselink and P. Adler]) established 16 new 1 m$^2$ plots located in one of the same exclosures as 6 of the original vegetation monitoring plots.
We avoided areas on steep hill slopes, areas with greater than 20\% cover of bare rock, and areas with greater than 10\% cover of the shrubs *Purshia tridentata* and/or *Amelanchier utahensis*.
We established the new plots in pairs and randomly assigned each plot in a pair to receive a "drought" or "irrigation" treatment.

@@

We monitored soil moisture in four of the drought-irrigation pairs using Decagon Devices (Pulman, Washington) 5TM and EC-5 soil moisture sensors.
We installed four sensors around the edges of each plot, two at 5 cm soil depth and two at 25 cm soil depth.
We also installed four sensors in areas nearby the four selected plot pairs to measure ambient soil moisture at the same depths.
Soil moisture measurements were automatically logged every four hours.
We coupled this temporally intensive soil moisture sampling with spatially extensive readings taken at six points within all 16 plots and associated ambient measurement areas.
These snapshot data were collected on 06/06/2012, 04/29/2015, 05/07/2015, 06/09/2015, and 05/10/2016 using a handheld EC-5 sensor.

Analyzing the response to experimental treatments was complicated by the fact that we did not directly monitor soil moisture in each plot on each day of the experiment.  
Only a subset of plots were equiped with soil moisture sensors, and within those plots, one or more of the sensors frequently failed to collect data. 
To remedy these problems, and to produce average daily soil moisture values for the ambient, drought, and irrigation conditions, we used a statistical model to predict the average treatment effects on soil moisture during the course of the experiment.

We first averaged the observed soil moisture for each day and within each plot. 
Then we standardized the averages within each plot group by subtracting the average ambient soil moisture in that plot group and dividing by the standard deviation of the ambient soil moisture in that plot group. 
We then found the difference between the standardized ambient soil moisture and the standardized drought and irrigation soil moisture within each plot group. 
These transformations ensured that the treatment effects in each plot were appropriately scaled by the local ambient conditions within each plot group.

We then modeled the daily deviation from ambient conditions of the drought and irrigation treatments using a linear mixed effects model with independent variables for treatment, season (winter, spring, summer, fall), rainfall, and all two-way interactions.  
Rainy days were defined as any day in which precipitation was recorded and average temperature was above 3$^{\circ}$C. 
The day immediately following rainfall was also classified as rainy. 
We fit the model using the 'lme4::lmer()' function [@Bates2015] in R [@R2016], with random effects for plot group and date.
We weighted observations by the number of unique sensors or spot measurements that were taken in each plot on that day. 
We then used the model to predict the average daily soil moisture in the treated plots based on the average daily ambient soil moisture.  
We could only predict soil moisture in the treated plots on days for which we took at least one ambient soil moisture measurement (Fig. 2B).

@@

\begin{figure}[!ht]
  \centering
      \includegraphics[height=7in]{../figures/Figure1.png}
  \caption{(A) Probability density of historical precipitation from 1926-2016, with the years of the experiment shown with arrows on the \emph{x}-axis. (B) Observed soil volumetric water content (VWC) over the course of the experiment. (C) Mean (points) ANPP and its standard deviation (error bars) for each year of the experiment.}
\end{figure}

### 2017-07-26 manuscript/usses_water_ms.tex
addresses ARK minor comments

% 
\begin{abstract}
Precipitation is predicted to become more variable in the western United
States, meaning years of above and below average precipitation will
become more common. Periods of drought and deluge are major drivers of
ecosystem functioning in water limited grasslands, but how ecosystems
respond to precipitation may change as the duration of above and below
average periods increases. Changes in ecosystem functional response
could reflect underlying changes in species composition or species
reaching physiological thresholds at extreme precipitation levels. Here
we report the results of a five-year experiment where we used drought
and irrigation treatments (approximately 50\% decrease/increase) to
investigate whether ecosystem functional response remains consistent at
extreme precipitation levels. We also use species-level adundance data
to link ecosystem functional response to community dynamics. Ecosystem
functional response, defined as the relationship between soil moisture
and aboveground net primary productivity, was suprisingly consistent
across drought, control, and irrigation treatments. Aboveground net
primary productivity (ANPP) responded positively to availabile soil
moisture, but the response was weak and did not vary across treatments.
There was also no evidence that treatment effects grew over time. The
similarity of ecosystem functional response across treatments was not
due to compensatory shifts at the plant community level, where species
composition among treatments was similar and remarkably stable over the
five years. At least in the short-term, ecosystem functional response
and community composition in this sagebrush steppe system is resistant
to increases and decreases in growing season precipitation, perhaps due
to bet-hedging strategies of constituent species.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

@@

At a given site, the functional response of ANPP to water availability
(e.g., soil moisture) can be characterized by fitting a model to
historical observations of ANPP and soil moisture. However, the fitted
functional response may provide an incomplete picture because future
conditions are likely to be outside the historical range of variability
\citep{Smith2011}. For example, historical trends may show a linear soil
moisture-ANPP relationship, but that relationship could potentially
saturate if soil moisture is pushed far beyond typical levels.
Saturating relationships are actually common \citep{Hsu2012}, perhaps
because other resources, like nitrogen, become more limiting in wet
years than dry years. Knowing the curvature of the soil moisture-ANPP
relationship is critical for understanding how ecosystems will respond
to chronic alterations in water availability.

@@

Manipulating potentially limiting resources offers a route to
understanding how ecosystems will respond to resource levels that fall
outside the historical range of variability
\citep{Avolio2015, Gherardi2015, Knapp2017}. Altering the amount of
precipitation over many years should provide insight into the time
scales at which water-limited ecosystems respond to chronic resource
alteration. Following the HRF, we propose four alternative predictions
for the effect of precipitation manipulation on the ecosystem functional
response to soil moisture, that is, the soil moisture-ANPP relationship
(Figure 1). The four predictions are based on possible outcomes at the
community (e.g., community composition) and ecosystem (e.g., soil
moisture-ANPP regression) levels.

@@

All four outcomes are possible in any given ecosystem, but the time
scales at which the different scenarios play out likely differ
\citep{Smith2009, Wilcox2016, Knapp2017}. Thus, our task is not to test
the validity of the HRF, but rather to amass information on how quickly
ecosystem functional responses change in different ecosystems. Likewise,
we need to understand whether changes at the ecosystem level are driven
community level changes or individual level responses.

@@

We conducted our precipitation manipulation experiment at the United
States Sheep Experimental Station (USSES) near Dubois, Idaho
(44.2\(^{\circ}\) N, 112.1\(^{\circ}\) W), 1500 m above sea level. The
vegetation is typical of high elevation sagebrush steppe. The plant
community is dominated by the shrub \emph{Artemesia tripartita} and
three perennial bunchgrasses, \emph{Pseudoroegneria spicata}, \emph{Poa
secunda}, and \emph{Hesperostipa comata}. During the period of our
experiment (2011 -- 2015), average mean annual precipitation was 265 mm
year\(\phantom{}^{-1}\) and mean monthly temperature ranged from
-5.2\(^{\circ}\)C in January to 21.8\(^{\circ}\)C in July.

@@

Between 1926 and 1932, range scientists at the USSES established 26
permanent 1 m\(^2\) quadrats to track vegetation change over time. In
2007, we (well, one of us {[}P. Adler{]}) relocated 14 of the original
quadrats, six of which were inside a large, permanent livestock
exclosure. We use these six plots as control plots that have recieved no
treatment, just ambient precipitation. In spring 2011, we (well, two of
us {[}A. Kleinhesselink and P. Adler{]}) established 16 new 1 m\(^2\)
plots. We avoided areas on steep hill slopes, areas with greater than
20\% cover of bare rock, and areas with greater than 10\% cover of the
shrubs \emph{Purshia tridentata} and/or \emph{Amelanchier utahensis}. We
established the new plots in pairs and randomly assigned each plot in a
pair to receive a ``drought'' or ``irrigation'' treatment.

@@

To make sure the treatments were having the desired effects, we
monitored soil moisture in four of the drought-irrigation pairs using
Decagon Devices (Pulman, Washington) 5TM and EC-5 soil moisture sensors.
We installed four sensors in each plot, two at 5 cm soil depth and two
at 25 cm soil depth. We also installed four sensors in areas nearby the
four selected plot pairs to measure ambient soil moisture at the same
depths. Soil moisture measurements were automatically logged every four
hours. We coupled this temporally intensive soil moisture sampling with
spatially extensive readings taken at six points within all 16 plots and
associated ambient measurement areas. These snapshot data were collected
on 06/06/2012, 04/29/2015, 05/07/2015, 06/09/2015, and 05/10/2016 using
a handheld EC-5 sensor.

\ark{ANDY: add paragraph on the statistical modeling of soil moisture to produce Figure 2B (same as your Figure 2 in the precip-demography-prediction dissertation chapter).}

@@

Third, the ecosystem and community resistance to drought and deluge we
observed could be a real phenomenon in this system, at least over the
five year time span we observed. The perennial plants in this cold
desert ecosystem are tolerant to drought conditions \citep[A.R.
Kleinhesselink, unpublished data]{Bazzaz1979, Franks2011}, meaning a
consistent ecosystem functional response at low soil moisture is
perfectly reasonable. For example, many of the perennial grasses in our
focal ecosystem avoid drought stress by growing early in the growing
season \citep[A.R. Kleinhesselink, personal observation]{Blaisdell1958}.
Likewise, the dominant shrub in our focal ecosystem, \emph{Artemisia
tripartita}, has access to water deep in the soil profile thanks to a
deep root system \citep{Germino2014}. As for the irrigation treatments,
the lack of a saturating response (e.g., a lower regression slope)
suggests water availability remained a limiting factor at the levels of
precipitation we were able to manipulate.

>>>>

% 
\begin{abstract}
Precipitation is predicted to become more variable in the western United
States, meaning years of above and below average precipitation will
become more common. Periods of drought and deluge are major drivers of
ecosystem functioning in water limited grasslands, but how ecosystems
respond to precipitation may change as the duration of above and below
average periods increases. Changes in ecosystem functional response
could reflect compensatory changes in species composition or species
reaching physiological thresholds at extreme precipitation levels. Here
we report the results of a five-year experiment where we used drought
and irrigation treatments (approximately 50\% decrease/increase) to
investigate whether ecosystem functional response remains consistent at
extreme precipitation levels. We also use species-level adundance data
to link ecosystem functional response to community dynamics. Ecosystem
functional response, defined as the relationship between soil moisture
and aboveground net primary productivity, was suprisingly consistent
across drought, control, and irrigation treatments. Aboveground net
primary productivity (ANPP) responded positively to availabile soil
moisture, but the response was weak and did not vary across treatments.
There was also no evidence that treatment effects grew over time. The
similarity of ecosystem functional response across treatments was not
due to compensatory shifts at the plant community level, where species
composition among treatments was similar and remarkably stable over the
five years. At least in the short-term, ecosystem functional response
and community composition in this sagebrush steppe system is consistent,
even at extreme precipitation levels. Such consistency may be due to
bet-hedging strategies of constituent species.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

@@

At a given site, the functional response of ANPP to water availability
(e.g., soil moisture) can be characterized by fitting a model to
historical observations of ANPP and soil moisture. However, the fitted
functional response may provide an incomplete picture because future
conditions are likely to be outside the historical range of variability
\citep{Smith2011}. For example, historical trends may show a linear soil
moisture-ANPP relationship, but that relationship could potentially
saturate if soil moisture is pushed far beyond typical levels.
Saturating relationships are actually common
\citep{Hsu2012, Gherardi2015a}, perhaps because other resources, like
nitrogen, become more limiting in wet years than dry years. Knowing the
curvature of the soil moisture-ANPP relationship is critical for
understanding how ecosystems will respond to chronic alterations in
water availability.

@@

Manipulating potentially limiting resources like precipitation offers a
route to understanding how ecosystems will respond to resource levels
that fall outside the historical range of variability
\citep{Avolio2015, Gherardi2015, Knapp2017}. Altering the amount of
precipitation over many years should provide insight into the time
scales at which water-limited ecosystems respond to chronic resource
alteration. Following the HRF, we propose four alternative predictions
for the effect of precipitation manipulation on the ecosystem functional
response to soil moisture, that is, the soil moisture-ANPP relationship
(Figure 1). The four predictions are based on possible outcomes at the
community (e.g., community composition) and ecosystem (e.g., soil
moisture-ANPP regression) levels.

@@

All four outcomes are possible in any given ecosystem, but the time
scales at which the different scenarios play out likely differ
\citep{Smith2009, Wilcox2016, Knapp2017}. Thus, our task is not to test
the validity of the HRF, but rather to amass information on how quickly
ecosystem functional responses change in different ecosystems. Likewise,
we need to understand whether changes at the ecosystem level are driven
by community level changes or individual level responses.

@@

We conducted our precipitation manipulation experiment in a sagebrush
steppe community at the United States Sheep Experimental Station (USSES)
near Dubois, Idaho (44.2\(^{\circ}\) N, 112.1\(^{\circ}\) W), 1500 m
above sea level. The plant community is dominated by the shrub
\emph{Artemesia tripartita} and three perennial bunchgrasses,
\emph{Pseudoroegneria spicata}, \emph{Poa secunda}, and
\emph{Hesperostipa comata}. During the period of our experiment (2011 --
2015), average mean annual precipitation was 265 mm
year\(\phantom{}^{-1}\) and mean monthly temperature ranged from
-5.2\(^{\circ}\)C in January to 21.8\(^{\circ}\)C in July.

@@

Between 1926 and 1932, range scientists at the USSES established 26
permanent 1 m\(^2\) quadrats to track vegetation change over time. In
2007, we (well, one of us {[}P. Adler{]}) relocated 14 of the original
quadrats, six of which were inside a large, permanent livestock
exclosure. We use these six plots as control plots that have recieved no
treatment, just ambient precipitation. In spring 2011, we (well, two of
us {[}A. Kleinhesselink and P. Adler{]}) established 16 new 1 m\(^2\)
plots located in one of the same exclosures as 6 of the original
vegetation monitoring plots. We avoided areas on steep hill slopes,
areas with greater than 20\% cover of bare rock, and areas with greater
than 10\% cover of the shrubs \emph{Purshia tridentata} and/or
\emph{Amelanchier utahensis}. We established the new plots in pairs and
randomly assigned each plot in a pair to receive a ``drought'' or
``irrigation'' treatment.

@@

We monitored soil moisture in four of the drought-irrigation pairs using
Decagon Devices (Pulman, Washington) 5TM and EC-5 soil moisture sensors.
We installed four sensors around the edges of each plot, two at 5 cm
soil depth and two at 25 cm soil depth. We also installed four sensors
in areas nearby the four selected plot pairs to measure ambient soil
moisture at the same depths. Soil moisture measurements were
automatically logged every four hours. We coupled this temporally
intensive soil moisture sampling with spatially extensive readings taken
at six points within all 16 plots and associated ambient measurement
areas. These snapshot data were collected on 06/06/2012, 04/29/2015,
05/07/2015, 06/09/2015, and 05/10/2016 using a handheld EC-5 sensor.

Analyzing the response to experimental treatments was complicated by the
fact that we did not directly monitor soil moisture in each plot on each
day of the experiment.\\
Only a subset of plots were equiped with soil moisture sensors, and
within those plots, one or more of the sensors frequently failed to
collect data. To remedy these problems, and to produce average daily
soil moisture values for the ambient, drought, and irrigation
conditions, we used a statistical model to predict the average treatment
effects on soil moisture during the course of the experiment.

We first averaged the observed soil moisture for each day and within
each plot. Then we standardized the averages within each plot group by
subtracting the average ambient soil moisture in that plot group and
dividing by the standard deviation of the ambient soil moisture in that
plot group. We then found the difference between the standardized
ambient soil moisture and the standardized drought and irrigation soil
moisture within each plot group. These transformations ensured that the
treatment effects in each plot were appropriately scaled by the local
ambient conditions within each plot group.

We then modeled the daily deviation from ambient conditions of the
drought and irrigation treatments using a linear mixed effects model
with independent variables for treatment, season (winter, spring,
summer, fall), rainfall, and all two-way interactions.\\
Rainy days were defined as any day in which precipitation was recorded
and average temperature was above 3\(^{\circ}\)C. The day immediately
following rainfall was also classified as rainy. We fit the model using
the `lme4::lmer()' function \citep{Bates2015} in R \citep{R2016}, with
random effects for plot group and date. We weighted observations by the
number of unique sensors or spot measurements that were taken in each
plot on that day. We then used the model to predict the average daily
soil moisture in the treated plots based on the average daily ambient
soil moisture.\\
We could only predict soil moisture in the treated plots on days for
which we took at least one ambient soil moisture measurement (Fig. 2B).

@@

Third, the ecosystem and community resistance to drought and deluge we
observed could be a real phenomenon in this system, at least over the
five year time span we observed. The perennial plants in this cold
desert ecosystem are tolerant to drought conditions \citep[A.R.
Kleinhesselink, unpublished data]{Bazzaz1979, Franks2011}, meaning a
consistent ecosystem functional response at low soil moisture is
perfectly reasonable. For example, many of the perennial grasses in our
focal ecosystem avoid drought stress by growing early in the growing
season \citep[A.R. Kleinhesselink, personal observation]{Blaisdell1958}.
Likewise, the dominant shrub in our focal ecosystem, \emph{Artemisia
tripartita}, has access to water deep in the soil profile thanks to a
deep root system (CITATION). As for the irrigation treatments, the lack
of a saturating response (e.g., a lower regression slope) suggests water
availability remained a limiting factor at the levels of precipitation
we were able to manipulate.
